---
title: "Weight construction"
author: "Sophia Müller-Dott"
date: '2022-09-22'
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(MatrixRider)
library(TFBSTools)
library(Biostrings)
library(tidyverse)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(BSgenome.Hsapiens.UCSC.hg38)
library('org.Hs.eg.db')
```


## Introduction

We want to add biolofical weights to our TF-gene connections. For that we will try to calculate binding affinities for each connection.
We need TF motifs and Promoter Sequence to calculate the binding affinities using MatrixRider.

These are all connections in CollecTRI
```{r}
GRN <- read.csv("output/040722/signed_CollecTRI_dbTF_040722.csv") %>%
  mutate(TF = map_chr(str_split(TF.TG, ":"), 1),
         TG = map_chr(str_split(TF.TG, ":"), 2))
```


## Promoter sequence
We select all target genes in our network and map their Gene Symbols to Entrez IDs. For seven Genes no entrez IDs are available
```{r}
symbols <- unique(GRN$TG)
# Gene symbols are transformed to entrez IDs
entrez_ids <- mapIds(org.Hs.eg.db, symbols, 'ENTREZID', 'ALIAS')
ids <- entrez_ids[!is.na(entrez_ids)]

names(entrez_ids)[is.na(entrez_ids)]
```


We then select 1000 nucleotides upstream of the transcription start site as the possible binding region of a TF.
The binding affinity of a TF to this sequence will be calculated. 
For 62 genes we have a problem accessing the promoter sequence

```{r}
transcriptCoordsByGene.GRangesList <- transcriptsBy(TxDb.Hsapiens.UCSC.hg38.knownGene, by = "gene")

prom_sequence <- map(unique(ids), function(id){
  getPromSeq <- function(id){
    tscripts_gr <- stack(transcriptCoordsByGene.GRangesList[id], "gene_id")
    res <- getPromoterSeq (tscripts_gr[which.min(tscripts_gr@ranges@start)],
                           Hsapiens, upstream=1000, downstream=100)
    return(res)
  }

  getPromSeq_poss <- possibly(.f = getPromSeq, otherwise = NA)
  getPromSeq_poss(id)

})
names(prom_sequence) <- unique(ids)
prom_sequence_raw <- prom_sequence
prom_sequence <- prom_sequence[!is.na(prom_sequence)]

mapIds(org.Hs.eg.db, names(prom_sequence_raw[is.na(prom_sequence_raw)]), 'ALIAS', 'ENTREZID') %>% unlist()

knitr::kable(data.frame(promoterSequence = length(prom_sequence),
                        noSequence = length(mapIds(org.Hs.eg.db, names(prom_sequence_raw[is.na(prom_sequence_raw)]), 'ALIAS', 'ENTREZID') %>% unlist())))
```


## TF motifs

For TF motifs we used JASPAR, hocomoco, unibind. 
encode is included in JASPAR
open questions:
- include mouse information to increase the coverage?
- How to handle multiple motifs in JASPAR. They might come from different sources, e.g. ChIP-Seq, Compiled(?). But I don't know how I can access that information in the files (only saw it on the website). 

### JASPAR
```{r}
JASPAR_down_core <- list.files("data/JASPAR2022_CORE_redundant_pfms_jaspar_20220923", full.names = T)
JASPAR_down_unval <- list.files("data/JASPAR2022_UNVALIDATED_redundant_pfms_jaspar_20220923", full.names = T)

TF_motifs <- map(c(JASPAR_down_core, JASPAR_down_unval), TFBSTools::readJASPARMatrix)

TFs <- map_chr(TF_motifs, function(tf){
  names(tf@listData)
})

humanTFs <- unique(GRN$TF)[unique(GRN$TF) %in% TFs]
notf <- unique(GRN$TF)[!unique(GRN$TF) %in% TFs] 

jaspar_tf_motifs <- map(TF_motifs, function(motif){
  motif@listData[[1]]
})

names(jaspar_tf_motifs) <- TFs
jaspar_tf_motifs <- jaspar_tf_motifs[!duplicated(names(jaspar_tf_motifs))]
jaspar_tf_motifs <- jaspar_tf_motifs[names(jaspar_tf_motifs) %in% unique(GRN$TF)]

knitr::kable(data.frame(motif = c("Human", "No"),
                        numberOfTFs = c(length(humanTFs), length(notf))))
```

### HOCOMOCO

HOCOMOCO stored information about 71 TFs which were not part of JASPAR. With Jaspar and HOCOMO we have the information of 646/839 TFs.
```{r}
hocomoco <- list.files("data/hocomoco_20220922", full.names = T)
hocomoco_tfs <- hocomoco[map_lgl(hocomoco, function(tf_file){
  tf <- str_split(str_split(tf_file, "_")[[1]][2], "/")[[1]][2]
  any(str_detect(notf, tf)) && notf[str_detect(notf, tf)] == tf
})]

hocomoco_tf_motifs <- map(hocomoco_tfs, function(tf_file){
  all.lines <- scan(tf_file, what=character(0), sep='\n', quiet=TRUE)
  pcm_raw <- map(2:length(all.lines), function(i){
     str_split(all.lines[i],  '\t') %>% unlist() %>% as.numeric()
  }) %>% unlist()
  
  profileMatrix <- matrix(as.integer(pcm_raw) ,nrow=4, byrow=TRUE, dimnames=list(DNA_BASES))

  pfmQuery <- PFMatrix(profileMatrix=profileMatrix, name = str_match(all.lines[1], ">\\s*(.*?)\\s*_")[2])
})

names(hocomoco_tf_motifs) <- map_chr(hocomoco_tfs, function(tf_file){
  all.lines <- scan(tf_file, what=character(0), sep='\n', quiet=TRUE)
  str_match(all.lines[1], ">\\s*(.*?)\\s*_")[2]
})
hocomoco_tf_motifs <- hocomoco_tf_motifs[!duplicated(names(hocomoco_tf_motifs))]

knitr::kable(data.frame(motif = c("Human", "No"),
                        numberOfTFs = c(sum(unique(GRN$TF) %in% unique(c(names(jaspar_tf_motifs), names(hocomoco_tf_motifs)))), 
                                        sum(!unique(GRN$TF) %in% unique(c(names(jaspar_tf_motifs), names(hocomoco_tf_motifs)))))))
```

### Unibind
In theory Unibind has 5 more TF but I don't understand the additional information
```{r}
still_missing_tfs <- unique(GRN$TF)[!unique(GRN$TF) %in% unique(c(names(jaspar_tf_motifs), names(hocomoco_tf_motifs)))]
files_unibind <- list.files("data/damo_hg38_PWMs_20220923", full.names = T)
unibind_tf_files <- files_unibind[map_lgl(files_unibind, function(tf_file){
  tf <- str_split(tf_file, "\\.")[[1]][3]
  any(str_detect(still_missing_tfs, tf)) && any(str_detect(tf, still_missing_tfs)) && still_missing_tfs[str_detect(still_missing_tfs, tf)] == tf
})]

unibind_tfs <- unique(map_chr(unibind_tf_files, function(x)str_split(x, "\\.")[[1]][3]))
still_missing_tfs <- still_missing_tfs[!still_missing_tfs %in% unibind_tfs]


#map(unibind_tf_files, function(folder){
#  print(folder)
#  
#  file_pwm <- list.files(folder, full.names = T)[1]
#  
#  
#  pwm <- scan(file_pwm, what=character(0), sep='\n', quiet=TRUE)
#
#})

```

### Mouse
If we would include binding sites from mice where we don't have any human information we would have additional informatio for 29 TFs (to a total of 679 TFs)
```{r}
tf_m <- paste0(substring(still_missing_tfs, 1, 1),
                 tolower(substring(still_missing_tfs, 2, nchar(still_missing_tfs))))

sum(tf_m %in% TFs)
```

### Check TFs without motif
```{r}
TFs_nm <- unique(GRN$TF)[!unique(GRN$TF) %in% unique(c(names(jaspar_tf_motifs),
                                                                  names(hocomoco_tf_motifs)))]

TF_size_nm <- GRN %>% 
  dplyr::filter(TF %in% TFs_nm) %>%
  group_by(TF) %>%
  summarise(n = n()) %>%
  arrange(desc(n))

ggplot(table(TF_size_nm$n) %>% as.data.frame(), aes(x = Var1, y= Freq)) +
  geom_bar(stat="identity") +
  xlab("Number of targets") +
  geom_vline(xintercept = 4.5, color = "red") +
  annotate("text", x=3, y=35, label= sum(TF_size_nm$n < 5), color = "red") +
  annotate("text", x=24, y=35, label= sum(TF_size_nm$n >= 5), color = "red")  + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

head(TF_size_nm)
```


## Calculate probabilities
### Matrix Rider

```{r}
final_TFmotifs <- append(jaspar_tf_motifs, hocomoco_tf_motifs)
motif_size <- data.frame(TF = names(final_TFmotifs),
                         size = map_dbl(final_TFmotifs, function(x){x@profileMatrix %>% ncol()}))

ggplot(table(motif_size$size) %>% as.data.frame(), aes(x = Var1, y= Freq)) +
  geom_bar(stat="identity") +
  xlab("Motif length") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


```{r}
GRN_prob <- map_dfr(1:nrow(GRN), function(j){
  print(j)
  x <- GRN[j,]
  if (any(names(ids) == x$TG)){
    gene_id <- ids[names(ids) == x$TG]
  } else {
    gene_id <- c("no gene id")
  }

  if (x$TF %in% names(final_TFmotifs) & gene_id %in% names(prom_sequence)){
    pfm <- final_TFmotifs[[x$TF]]
    motif_length <- pfm@profileMatrix %>% ncol()
    sequ <- prom_sequence[[gene_id]][[1]]
    bind_aff <- log(getSeqOccupancy(sequ, pfm, 0), base = 10)

    data.frame(TF = x$TF,
               TG = x$TG,
               bind_aff = bind_aff,
               motif_length = motif_length)
  } else {
    data.frame(TF = x$TF,
               TG = x$TG,
               bind_aff = NA,
               motif_length = NA)
  }
})

GRN_prob$bind_aff[is.infinite(GRN_prob$bind_aff)] <- NA

saveRDS(GRN_prob, "output/040722/binding_affinities_matrixRider_dbTF_new.rds")
```

```{r}
GRN_prob_filtered <- GRN_prob %>%
  filter(!is.na(bind_aff))
ggplot(GRN_prob_filtered, aes(x = bind_aff, y = motif_length)) +
  geom_point()
cor(GRN_prob_filtered$bind_aff, GRN_prob_filtered$motif_length)

GRN_prob_filtered <- GRN_prob_filtered %>%
  mutate(bind_aff_length = bind_aff/(1100-motif_length+1))

ggplot(GRN_prob_filtered, aes(x = bind_aff_length, y = motif_length)) +
  geom_point()
cor(GRN_prob_filtered$bind_aff_length, GRN_prob_filtered$motif_length)

motif_score <- GRN_prob_filtered %>% group_by(motif_length) %>% summarize(mean_bind = mean(bind_aff))

p <- ggplot(motif_score, aes(x = mean_bind, y = motif_length)) +
  geom_point() +
  theme_minimal() + theme(text = element_text(size = 20)) +
  xlab("mean binding affinity")

pdf("figures/weight_distribution_TF.pdf", width = 7, height = 6)
p
dev.off()
cor(motif_score$mean_bind, motif_score$motif_length)
```


### Scaling
Edges where either the TF motif or the sequence and all edges with a binding score lower than zero were removed. The weights were then scaled to be between 0 and 1. 
```{r}
GRN_prob <- readRDS("output/040722/binding_affinities_matrixRider_dbTF_new.rds")

TF_weights <- GRN_prob %>%
  filter(!is.na(bind_aff.mean)) %>%
  group_by(TF) %>%
  summarize(std_weight = sd(bind_aff.mean))

TF_weights <- GRN_prob %>%
  filter(!is.na(bind_aff)) %>%
  group_by(TF) %>%
  summarize(mean_weight = mean(bind_aff))

# Distribution of weights
dist_weight_p <- ggplot(GRN_prob,aes(x = bind_aff.mean)) +
  geom_density() +
  theme_minimal() + theme(text = element_text(size = 20)) +
  xlab("binding affinity")

pdf("figures/weight_distribution.pdf", width = 9, height = 6)
dist_weight_p
dev.off()

dist_weight_tf_p <- ggplot(TF_weights,aes(x = std_weight)) + geom_density()

# Remove edges with binding probability of zero or lower
GRN_prob_filter <- GRN_prob %>%
  filter(!is.na(bind_aff)) %>%
  mutate(bind_aff = bind_aff + abs(min(bind_aff)))

ggplot(GRN_prob_filter,aes(x = bind_aff)) +
  geom_density() +
  theme_minimal() + theme(text = element_text(size = 20))
```

## scale per gene
```{r}
GRN_scaled <- map_dfr(unique(GRN_prob_filter$TG), function(tg){
  GRN_tf <- GRN_prob_filter %>% dplyr::filter(TG == tg)

  GRN_tf_tmp <- GRN_tf %>% mutate(scaled_prob = (GRN_tf$bind_aff) / (max(GRN_tf$bind_aff, na.rm = T)))
  GRN_tf_tmp %>% dplyr::select(TF, TG, scaled_prob)
})

GRN_scaled <- GRN_scaled %>%
  left_join(GRN_prob_filter, by = c("TF", "TG"))

#cut_off <- quantile(GRN_scaled$bind_aff, probs = 0.2)
#filter <- GRN_scaled$TG[GRN_scaled$scaled_prob == 1 & GRN_scaled$bind_aff <= cut_off]

#GRN_scaled <- GRN_scaled %>%
#  filter(!TG %in% filter)
```


```{r}
GRN_scaled <- map_dfr(unique(GRN_prob_filter$TF), function(tf){
  GRN_tf <- GRN_prob_filter %>% dplyr::filter(TF == tf)

  GRN_tf_tmp <- GRN_tf %>% mutate(scaled_prob = (GRN_tf$bind_aff.mean) / (max(GRN_tf$bind_aff.mean, na.rm = T)))
  GRN_tf_tmp %>% dplyr::select(TF, TG, scaled_prob)
})
```


```{r}
motif_length_df <- GRN_scaled %>% 
  group_by(motif_length) %>%
  summarize(mean_bind = mean(scaled_prob))

cor(motif_length_df$motif_length, motif_length_df$mean_bind)

ggplot(motif_length_df, aes(x = motif_length, y= mean_bind)) +
  geom_point()

dist_weight_sclaed_p <- ggplot(GRN_scaled,aes(x = scaled_prob)) +
  geom_density()

TFs <- table(GRN_scaled$TF) %>% as.data.frame() %>% filter(Freq >= 5) %>% pull(Var1)
dist_weight_scaled_tf_p <- ggplot(GRN_scaled %>%
                                    filter(TF %in% TFs) %>%
                                    group_by(TF) %>%
                                    summarize(mean_weight = mean(scaled_prob)),aes(x = mean_weight)) + geom_density()
```


```{r}
GRN_final <- GRN_scaled %>% 
  mutate(TF.TG = paste(TF, TG, sep = ":")) %>%
  dplyr::select(TF.TG, scaled_prob) %>%
  left_join(GRN, by = "TF.TG") %>%
  mutate(weight = scaled_prob * sign) %>%
  dplyr::select(TF, TG, weight, PMID) %>%
  rename("source" = "TF", "target" = "TG") %>%
  arrange(source, target)

test <- GRN_final %>% group_by(source) %>%
  summarize(mean_weiht = mean(abs(weight)),
            n = n(),
            sum_weight = sum(abs(weight)))

test_2 <- test %>% 
  group_by(n) %>%
  summarize(mean_weiht =  mean(abs(mean_weiht)))

ggplot(test_2, aes(x=mean_weiht, y=n)) +
  geom_point()

ggplot(test_2, aes(x=mean_weiht, y=log(n))) +
  geom_point()

cor(test$mean_weiht, test$n, method = "spearman")
head(GRN_final)

write_csv(GRN_final, "output/040722/GRN_weighted_gene_dbTF_040722.csv")
```


### Coverage
```{r}
GRN_missingInfo <- GRN_prob %>% filter(!is.na(bind_aff.mean))
GRN_size <- data.frame(GRN_type = rep(c("GRN signed", "GRN missing information", "GRN weighted"), each = 4),
           type = rep(c("Edges", "TF", "TG", "Mean Regulon size"), times = 3),
           value = c(nrow(GRN), length(unique(GRN$TF)), length(unique(GRN$TG)), mean(table(GRN$TF)),
                     nrow(GRN_missingInfo), length(unique(GRN_missingInfo$TF)), length(unique(GRN_missingInfo$TG)), mean(table(GRN_missingInfo$TF)),
                     nrow(GRN_final), length(unique(GRN_final$source)), length(unique(GRN_final$target)), mean(table(GRN_final$source))
                     )
           )

ggplot(GRN_size) + 
  aes(x = type, y = value, fill = GRN_type) +
  geom_bar(stat="identity", position=position_dodge())+
  ggbreak::scale_y_cut(breaks=c(100, 1000, 10000), which=c(1,1,1 ), scales=c(0.5, 0.3, 0.1)) +
  theme_minimal() + xlab("") + ylab("")
  
# Distribution of scaledweights
dist_weight_scaled_p <- ggplot(GRN_final,aes(x = weight)) +
  geom_density()

TF_weights_scaled <- GRN_final %>%
  group_by(source) %>%
  summarise(mean_weight = mean(weight, na.rm = T)) %>%
  arrange(desc(mean_weight))


dist_weight_scaled_tf_p <- ggplot(TF_weights_scaled,aes(x = mean_weight)) + geom_density()
dist_weight_scaled_p
dist_weight_scaled_tf_p
```


## Probabilities for filtering
```{r}
GRN_tmp <- GRN_prob %>%
  filter(!is.na(bind_aff.mean))

quan_10 <- quantile(GRN_tmp$bind_aff.mean, probs = c(0.10), na.rm = T)
quan_20 <- quantile(GRN_tmp$bind_aff.mean, probs = c(0.20), na.rm = T)
quan_30 <- quantile(GRN_tmp$bind_aff.mean, probs = c(0.30), na.rm = T)
quan_40 <- quantile(GRN_tmp$bind_aff.mean, probs = c(0.40), na.rm = T)
quan_50 <- quantile(GRN_tmp$bind_aff.mean, probs = c(0.50), na.rm = T)

quantiles <- c(quan_10, quan_20, quan_30, quan_40, quan_50)
map(names(quantiles), function(q_n){
  q <- quantiles[q_n]
  GRN_quant <- GRN_tmp %>%
  filter(bind_aff.mean >= q) %>%
  left_join(GRN, by = c("TF", "TG")) %>%
  rename(c("weight" = "sign", "source" = "TF", "target" = "TG")) %>%
  mutate(confidence = "A") %>%
  dplyr::select(c(source, confidence, target, weight))
  
  write_csv(GRN_quant, file = paste0("output/040722/filter_dbTF/collecTRI_", q_n, ".csv"))
  
  edges_rm <- sum(!GRN_tmp$bind_aff.mean >= q)
  GRN_quant_random <- GRN_tmp[-sample(1:nrow(GRN_tmp), edges_rm), ] %>%
  left_join(GRN, by = c("TF", "TG")) %>%
  rename(c("weight" = "sign", "source" = "TF", "target" = "TG")) %>%
  mutate(confidence = "A") %>%
  dplyr::select(c(source, confidence, target, weight))
  
  write_csv(GRN_quant_random, file = paste0("output/040722/filter_dbTF/collecTRI_", q_n, "_random.csv"))
})

```
